---
title: "IPUMS Data aggregation"
author: "David Gill and Matheus De Nardo"
date: "September 11, 2017"
output: html_document
---

## IPUMS data
This code is used to aggregate social and economic variables from IPUMS at the Geolevel 1 and 2 scale. These data will be used to assess social and economic differences in MPA vs non-MPA communities, both in terms of placement and changes over time. Here data are derived from the IPUMS website on educational attainment and employment, as well as other demographic and economic variables.
### Read in R packages
```{r eval=FALSE}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(data.table)
#library(ff)
#library(ffbase2)
library(rgdal)
options(scipen=999,stringsAsFactors = F)
sum2=function(x){sum(x,na.rm=TRUE)}
mean2=function(x){mean(x,na.rm=TRUE)}
memory.limit(size = 32000) 

# location of downloaded microdata (country csvs)
loc_files <- 'E:/My Documents/Important Files/Work/SESYNC/Data analysis/IPUMS/data/ipumsi_countries/'
file.list<- list.files(loc_files,recursive = T, full.names = F) # get list of files in folder

# location of shapefiles
dsn <- "E:/My Documents/Important Files/Work/SESYNC/GIS/IPUMS"# location of shapefiles
```

## Steps
1. change the file locations above, placing the country csvs in your *loc_files* folder
2. run the two chunks to read in the 2 functions listed below
3. for the section "Compile all regions" run line by line. *IPUMS_aggr(1:length(file.list))* will run all at once, *IPUMS_aggr(1)* will just run it for the 1st csv. Running all of them will take some time (~2 hours for all 82 countries)
4. group the data (e.g. average by decade)
5. join the data to the shapefiles in the *dsn* folder


## Summing and aggregation functions
There are two functions:

**1. IPUMS_summary**
This function does the following:

* aggregates the data for the select variables of interest at the geolevel2 scale. 
* writes the aggregated data to a new file called **sum.table** with the summarized values and **pct.table** with the percentages

**2. IPUMS_aggr**
This function does the following:

* finds and loops through all the files in the folder with the countries
* calculates the percentages

*(need to clean this and place this somewhere else)*
The variables of interest are:
* **Total population**: this gives us total population (based on person weights) within the geography of interest (*popXX*) 
* **Education**: the % of the population who are 25+ years old, for whom educ. attainment is determined, with secondary (*phsXX*) or university (*pcolXX*) level education
* **Unemployment**: the % of individuals within the total workforce (excluding inactive members), for whom employment status is determined, who are unemployed  (*punempXX*).
* **Professional**: the % of individuals within the total employed workforce, for whom employment status is determined, who are employed in a professional occupation  (*pprofXX*). 
* **Female labor force employment**: the % of individuals within the total employed workforce, for whom employment status is determined, who are employed in a professional occupation  (*pflabfXX*).  


### IPUMS_summary function 
summarizes raw values
```{r eval=FALSE}
IPUMS_summary <- function (x) {
  join.names <- c("COUNTRY", "YEAR", "GEOLEV1", "GEOLEV2")
  x <- filter(x,!is.na(GEOLEV1))   # remove entries without Geolevel data (shouldn't happen!)

if (nrow(x)>0 & nrow(filter(x,is.na(AGE)))/nrow(x)<0.1) {  # run if countries are not missing most of the data

tot.popn <- x %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(popXX = sum(PERWT))
 
#--Education                                    
# number of persons with hs degree or less (Secondary education and post-secondary technical)
  if (class(x$EDATTAIND)!="NULL" & !all(is.na(x$EDATTAIND))) {
educsec <- filter(x,EDATTAIND%in%(311:322) & AGE%in%c(25:100)) %>% # ages 25+ yrs
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(edusecondXX = sum(PERWT))
educD <- filter(x,EDATTAIND%in%(100:400) & AGE%in%c(25:100)) %>% # ages 25+ yrs
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(educDXX = sum(PERWT))

# number persons with at least a four year college degree
educter <- filter(x,EDATTAIND==400 & AGE%in%c(25:100)) %>% # ages 25+ yrs
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(educterXX = sum(PERWT))
   }
 # if EDATTAIND is missing but EDATTAIN is present
#  else if (class(x$EDATTAIN)!="NULL" & all(!is.na(x$EDATTAIN))) {
#educsec <- filter(x,EDATTAIN%in%(3) & AGE%in%c(25:100)) %>% # ages 25+ yrs
#  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
#  summarize(edusecondXX = sum(PERWT))
#educD <- filter(x,EDATTAIN%in%(1:4) & AGE%in%c(25:100)) %>% # ages 25+ yrs
#  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
#  summarize(educDXX = sum(PERWT))

# number persons with at least a four year college degree
#educter <- filter(x,EDATTAIN==4 & AGE%in%c(25:100)) %>% # ages 25+ yrs
#  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
#  summarize(educterXX = sum(PERWT))
#   }
  else{
  print(paste0("EDATTAIND is either missing or blank for country ",x$COUNTRY[1]))
  educsec <- data.frame(tot.popn[,1:4],edusecondXX=NA);educD <- data.frame(tot.popn[,1:4],educDXX=NA);
  educter <- data.frame(tot.popn[,1:4],educterXX=NA);
  }
  

# Percent of persons age 10+ literate(numerator)"
  if (class(x$LIT)!="NULL"& !all(is.na(x$LIT))) {
lit10N <- filter(x,LIT==2 & AGE%in%c(10:100)) %>% # ages 25+ yrs
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(lit10NXX = sum(PERWT))

# Percent of persons age 10+ literate (denominator)"
lit10D <- filter(x,LIT%in%c(1:2) & AGE%in%c(10:100)) %>% # ages 25+ yrs
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(lit10DXX = sum(PERWT))
   }
  else{
  lit10N <- data.frame(tot.popn[,1:4],lit10NXX=NA);lit10D <- data.frame(tot.popn[,1:4],lit10DXX=NA)
    }

educ <- tot.popn[join.names] %>% 
  left_join(educsec,by = join.names) %>% 
  left_join(educter,by = join.names) %>% 
  left_join(educD,by = join.names) %>% 
  left_join(lit10N,by = join.names) %>% 
  left_join(lit10D,by = join.names)

#--Employment
# Current labor force (employed and unemployed individuals between 16-65; excludes housework)
#emptlf <- filter(x,EMPSTATD%in%c(100:240) & AGE%in%c(16:65)) %>%
#  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
#  summarize(emptlfXX = sum(PERWT))

# Total labor force- persons 16 years and over
  if (class(x$EMPSTATD)!="NULL" & !all(is.na(x$EMPSTATD))) {
tlf <- filter(x,EMPSTATD%in%c(100:240) & AGE%in%c(16:65)) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>% #eliminating other column headers
  summarize(tlfXX = sum(PERWT)) #can add abbreviated function due to <%> piping
# Unemployment for individuals between 16-65 (of civilians)
unempl <- filter(x,EMPSTATD%in%c(200:240) & AGE%in%c(16:65)) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(unempXX = sum(PERWT)) 

# Females 16 and over, not in armed forces (generally, employed or actively seeking employment)
lfpfemN <- filter(x,SEX==2 & EMPSTATD%in%c(100:140) & AGE%in%c(16:65)) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(lfpfemNXX= sum(PERWT))
lfpfemD <- filter(x,SEX==2 & EMPSTATD%in%c(100:240) & AGE%in%c(16:65)) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(lfpfemDXX= sum(PERWT))
   }
  else{
  print(paste0("EMPSTATD is either missing or blank for country ",x$COUNTRY[1]))
  tlf <- data.frame(tot.popn[,1:4],tlfXX=NA);unempl <- data.frame(tot.popn[,1:4],unempXX=NA);
  lfpfemN <- data.frame(tot.popn[,1:4],lfpfemNXX=NA);lfpfemD <- data.frame(tot.popn[,1:4],lfpfemDXX=NA);
    }
# Percent of workers in professional or technical occupations (excluding clerks)
  if (class(x$EMPSTATD)!="NULL" & !all(is.na(x$EMPSTATD)) & class(x$OCCISCO)!="NULL" & !all(is.na(x$OCCISCO))) {
occprofN <- filter(x,AGE%in%c(16:65) & EMPSTATD%in%c(100:140) & OCCISCO%in%c(1:3)) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(occprofNXX = sum(PERWT)) 
occprofD <- filter(x,AGE%in%c(16:65) & EMPSTATD%in%c(100:140) & !OCCISCO%in%c(10,97:99)) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(occprofDXX = sum(PERWT)) 
  }
  else{
  print(paste0("OCCISCO are either missing or blank for country ",x$COUNTRY[1]))
  occprofN <- data.frame(tot.popn[,1:4],occprofNXX=NA);occprofD <- data.frame(tot.popn[,1:4],occprofDXX=NA)
  }

# Percent of currently employed persons age 16 to 65 who are working in agriculture or fisheries
  if (class(x$EMPSTATD)!="NULL" & !all(is.na(x$EMPSTATD)) &  class(x$INDGEN)!="NULL" & !all(is.na(x$INDGEN))) {
occagrN <- filter(x,AGE%in%c(16:65) & EMPSTATD%in%c(100:140) & INDGEN==10) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(occagrNXX = sum(PERWT)) 
occagrD <- filter(x,AGE%in%c(16:65) & EMPSTATD%in%c(100:140) & !INDGEN%in%c(0,998:999)) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(occagrDXX = sum(PERWT)) 
  }
  else{
  print(paste0("INDGEN are either missing or blank for country ",x$COUNTRY[1]))
  occagrN <- data.frame(tot.popn[,1:4],occagrNXX=NA);occagrD <- data.frame(tot.popn[,1:4],occagrDXX=NA)
  }

empl <- tot.popn[join.names] %>%
  left_join(tlf,by=join.names) %>%
  left_join(unempl,by=join.names) %>%
  left_join(lfpfemN,by=join.names) %>%
  left_join(lfpfemD,by=join.names) %>%
  left_join(occprofN,by=join.names) %>%
  left_join(occprofD,by=join.names) %>%
  left_join(occagrN,by=join.names) %>%
  left_join(occagrD,by=join.names) 


#--Assets
# Refrigerator ownership
  if (class(x$REFRIG)!="NULL" & !all(is.na(x$REFRIG))) {
refrigN <- filter(x,REFRIG==2 & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(refrigNXX = sum(HHWT))
refrigD <- filter(x,REFRIG%in%c(1:2) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(refrigDXX = sum(HHWT)) 
   }
  else{
  refrigN <- data.frame(tot.popn[,1:4],refrigNXX=NA);refrigD <- data.frame(tot.popn[,1:4],refrigDXX=NA)
    }

# hot water access (water heater in unit)
  if (class(x$HOTWATER)!="NULL" & !all(is.na(x$HOTWATER))) {
hotwaterN <- filter(x,HOTWATER==2 & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(hotwaterNXX = sum(HHWT))
hotwaterD <- filter(x,HOTWATER%in%c(1:2) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(hotwaterDXX = sum(HHWT)) 
   }
  else{
  hotwaterN <- data.frame(tot.popn[,1:4],hotwaterNXX=NA);hotwaterD <- data.frame(tot.popn[,1:4],hotwaterDXX=NA)
    }

# cellphone access
  if (class(x$CELL)!="NULL" & !all(is.na(x$CELL))) {
  cellN <- filter(x,CELL==1 & GQ==10 & PERNUM==1) %>%
        group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
        summarize(cellNXX = sum(HHWT))
  cellD <- filter(x,CELL%in%c(1:2) & GQ==10 & PERNUM==1) %>%
        group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
        summarize(cellDXX = sum(HHWT))
   }
  else{
  cellN <- data.frame(tot.popn[,1:4],cellNXX=NA);cellD <- data.frame(tot.popn[,1:4],cellDXX=NA)
    }


# TV ownership
  if (class(x$TV)!="NULL" & !all(is.na(x$TV))) {
tvN <- filter(x,TV%in%c(20:43) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(tvNXX = sum(HHWT))
tvD <- filter(x,TV%in%c(10:43) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(tvDXX = sum(HHWT)) 
   }
  else{
  tvN <- data.frame(tot.popn[,1:4],tvNXX=NA);tvD <- data.frame(tot.popn[,1:4],tvDXX=NA)
    }

# Radio ownership
  if (class(x$RADIO)!="NULL" & !all(is.na(x$RADIO))) {
radioN <- filter(x,RADIO==2 & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(radioNXX = sum(HHWT))
radioD <- filter(x,RADIO%in%c(1:2) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(radioDXX = sum(HHWT)) 
   }
  else{
  radioN <- data.frame(tot.popn[,1:4],radioNXX=NA);radioD <- data.frame(tot.popn[,1:4],radioDXX=NA)
    }

# phone ownership
  if (class(x$PHONE)!="NULL" & !all(is.na(x$PHONE))) {
phoneN <- filter(x,PHONE==2 & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(phoneNXX = sum(HHWT))
phoneD <- filter(x,PHONE%in%c(1:2) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(phoneDXX = sum(HHWT)) 
   }
  else{
  phoneN <- data.frame(tot.popn[,1:4],phoneNXX=NA);phoneD <- data.frame(tot.popn[,1:4],phoneDXX=NA)
    }

# computer ownership
  if (class(x$COMPUTER)!="NULL" & !all(is.na(x$COMPUTER))) {
computerN <- filter(x,COMPUTER==2 & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(computerNXX = sum(HHWT))
computerD <- filter(x,COMPUTER%in%c(1:2) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(computerDXX = sum(HHWT)) 
   }
  else{
  computerN <- data.frame(tot.popn[,1:4],computerNXX=NA);computerD <- data.frame(tot.popn[,1:4],computerDXX=NA)
    }

# internet access
  if (class(x$INTERNET)!="NULL" & !all(is.na(x$INTERNET))) {
internetN <- filter(x,INTERNET==2 & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(internetNXX = sum(HHWT))
internetD <- filter(x,INTERNET%in%c(1:2) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(internetDXX = sum(HHWT)) 
   }
  else{
  internetN <- data.frame(tot.popn[,1:4],internetNXX=NA);internetD <- data.frame(tot.popn[,1:4],internetDXX=NA)
    }

# Washer ownership
  if (class(x$WASHER)!="NULL" & !all(is.na(x$WASHER))) {
washerN <- filter(x,WASHER%in%c(2:4) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(washerNXX = sum(HHWT))
washerD <- filter(x,WASHER%in%c(1:4) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(washerDXX = sum(HHWT)) 
   }
  else{
  washerN <- data.frame(tot.popn[,1:4],washerNXX=NA);washerD <- data.frame(tot.popn[,1:4],washerDXX=NA)
    }

# Air-conditioner ownership
  if (class(x$AIRCON)!="NULL" & !all(is.na(x$AIRCON))) {
airconN <- filter(x,AIRCON%in%c(20:29) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(airconNXX = sum(HHWT))
airconD <- filter(x,AIRCON%in%c(10:29) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(airconDXX = sum(HHWT)) 
   }
  else{
  airconN <- data.frame(tot.popn[,1:4],airconNXX=NA);airconD <- data.frame(tot.popn[,1:4],airconDXX=NA)
    }

# Auto ownership
  if (class(x$AUTOS)!="NULL" & !all(is.na(x$AUTOS))) {
autoN <- filter(x,AUTOS%in%c(1:7) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(autoNXX = sum(HHWT))
autoD <- filter(x,AUTOS%in%c(0:7) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(autoDXX = sum(HHWT)) 
   }
  else{
  autoN <- data.frame(tot.popn[,1:4],autoNXX=NA);autoD <- data.frame(tot.popn[,1:4],autoDXX=NA)
    }


assets <- tot.popn[join.names]   %>%
  left_join(hotwaterN,by=join.names) %>%
  left_join(hotwaterD,by=join.names) %>%
  left_join(cellN,by=join.names) %>%
  left_join(cellD,by=join.names) %>%
  left_join(refrigN,by=join.names) %>%
  left_join(refrigD,by=join.names) %>%
  left_join(tvN,by=join.names) %>%
  left_join(tvD,by=join.names) %>% 
  left_join(radioN,by=join.names) %>%
  left_join(radioD,by=join.names) %>%
  left_join(phoneN,by=join.names) %>%
  left_join(phoneD,by=join.names) %>%
  left_join(computerN,by=join.names) %>%
  left_join(computerD,by=join.names) %>%
  left_join(internetN,by=join.names) %>%
  left_join(internetD,by=join.names) %>%
  left_join(washerN,by=join.names) %>%
  left_join(washerD,by=join.names) %>%
  left_join(airconN,by=join.names) %>%
  left_join(airconD,by=join.names) %>%
  left_join(autoN,by=join.names) %>%
  left_join(autoD,by=join.names)


#-- Household attributes
#Percent of households with electricity
  if (class(x$ELECTRIC)!="NULL" & !all(is.na(x$ELECTRIC))) {
electricN <- filter(x,ELECTRIC==1 & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(electricNXX = sum(HHWT))
electricD <- filter(x,ELECTRIC%in%c(1:2) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(electricDXX = sum(HHWT)) 
   }
  else{
  electricN <- data.frame(tot.popn[,1:4],electricNXX=NA);electricD <- data.frame(tot.popn[,1:4],electricDXX=NA)
    }

#Percent of households with a flush toilet
  if (class(x$TOILET)!="NULL" & !all(is.na(x$TOILET))) {
flushtoiletN <- filter(x,TOILET==21 & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize( flushtoiletNXX = sum(HHWT))
flushtoiletD <- filter(x,TOILET%in%c(10:23) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(flushtoiletDXX = sum(HHWT))
   }
  else{
  flushtoiletN <- data.frame(tot.popn[,1:4],flushtoiletNXX=NA);flushtoiletD <- data.frame(tot.popn[,1:4],flushtoiletDXX=NA)
    }

# Cooking fuel
#Electricity
  if (class(x$FUELCOOK)!="NULL" & !all(is.na(x$FUELCOOK))) {
cookelectN <- filter(x,FUELCOOK==20 & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(cookelectNXX = sum(HHWT))
cookD <- filter(x,FUELCOOK%in%c(10:77) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(cookDXX = sum(HHWT))
#Gas
cookgasN <- filter(x,FUELCOOK%in%c(30:47) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(cookgasNXX = sum(HHWT))
#Wood and solid fuels
cookwoodN <- filter(x,FUELCOOK%in%c(50:56) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(cookwoodNXX = sum(HHWT))
   }
  else{
  cookelectN <- data.frame(tot.popn[,1:4],cookelectNXX=NA); cookD <- data.frame(tot.popn[,1:4],cookDXX=NA);
  cookgasN <- data.frame(tot.popn[,1:4],cookgasNXX=NA); cookwoodN <- data.frame(tot.popn[,1:4],cookwoodNXX=NA)
    }

#Percentage of households with access to piped water
  if (class(x$WATSUP)!="NULL" & !all(is.na(x$WATSUP))) {
pipedN <- filter(x,WATSUP%in%c(10:18) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(pipedNXX = sum(HHWT))
pipedD <- filter(x,WATSUP%in%c(10:20) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(pipedDXX = sum(HHWT))
   }
  else{
  pipedN <- data.frame(tot.popn[,1:4],pipedNXX=NA);pipedD <- data.frame(tot.popn[,1:4],pipedDXX=NA)
    }

#Percentage of households with access to sewage facilities
  if (class(x$SEWAGE)!="NULL" & !all(is.na(x$SEWAGE))) {
sewageN <- filter(x,SEWAGE%in%c(10:12) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(sewageNXX = sum(HHWT))
sewageD <- filter(x,SEWAGE%in%c(10:20) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(sewageDXX = sum(HHWT))
   }
  else{
  sewageN <- data.frame(tot.popn[,1:4],sewageNXX=NA);sewageD <- data.frame(tot.popn[,1:4],sewageDXX=NA)
    }

#Percentage household ownership
  if (class(x$OWNERSHIP)!="NULL" & !all(is.na(x$OWNERSHIP))) {
homeownN <- filter(x,OWNERSHIP==1 & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(homeownNXX = sum(HHWT))
homeownD <- filter(x,OWNERSHIP%in%c(1:2) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(homeownDXX = sum(HHWT))
   }
  else{
  homeownN <- data.frame(tot.popn[,1:4],homeownNXX=NA);homeownD <- data.frame(tot.popn[,1:4],homeownDXX=NA)
    }

#Percentage land ownership
  if (class(x$LANDOWN)!="NULL" & !all(is.na(x$LANDOWN))) {
landownN <- filter(x,LANDOWN%in%c(10:14) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(landownNXX = sum(HHWT))
landownD <- filter(x,LANDOWN%in%c(10:30) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(landownDXX = sum(HHWT))
   }
  else{
  landownN <- data.frame(tot.popn[,1:4],landownNXX=NA);landownD <- data.frame(tot.popn[,1:4],landownDXX=NA)
    }

#Roof types
# Masonary, concrete, tiles
  if (class(x$ROOF)!="NULL" & !all(is.na(x$ROOF))) {
roofmasonN <- filter(x,ROOF%in%c(10:29) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(roofmasonNXX = sum(HHWT))
roofD <- filter(x,ROOF%in%c(10:90) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(roofDXX = sum(HHWT))
# Metal
roofmetalN <- filter(x,ROOF%in%c(30:35) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(roofmetalNXX = sum(HHWT))
# Wood and/or clay
roofwoodN <- filter(x,ROOF%in%c(40:61) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(roofwoodNXX = sum(HHWT))
   }
  else{
  roofmasonN <- data.frame(tot.popn[,1:4],roofmasonNXX=NA);roofD <- data.frame(tot.popn[,1:4],roofDXX=NA);
  roofmetalN <- data.frame(tot.popn[,1:4],roofmetalNXX=NA);roofwoodN <- data.frame(tot.popn[,1:4],roofwoodNXX=NA)
    }

#trash
# Collected or not collected
  if (class(x$TRASH)!="NULL" & !all(is.na(x$TRASH))) {
trashcollectN <- filter(x,TRASH%in%c(10:14) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(trashcollectNXX = sum(HHWT))
trashcollectD <- filter(x,TRASH%in%c(10:39) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(trashcollectDXX = sum(HHWT))
   }
  else{
  trashcollectN <- data.frame(tot.popn[,1:4],trashcollectNXX=NA);trashcollectD <- data.frame(tot.popn[,1:4],trashcollectDXX=NA)
    }

#Person per household                                      
hhsizeN <- filter(x, GQ==10 & PERNUM==1) %>% 
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(hhsizeNXX = sum2(PERSONS*HHWT))
hhsizeD <- filter(x, GQ==10 & PERNUM==1) %>% 
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(hhsizeDXX = sum2(HHWT))

houseatt <- tot.popn[join.names]   %>%
  left_join(electricN,by=join.names) %>%
  left_join(electricD,by=join.names) %>%
  left_join(flushtoiletN,by=join.names) %>%
  left_join(flushtoiletD,by=join.names) %>%
  left_join(cookelectN,by=join.names) %>%
  left_join(cookgasN,by=join.names) %>%
  left_join(cookwoodN,by=join.names) %>%
  left_join(cookD,by=join.names) %>% 
  left_join(pipedN,by=join.names) %>%
  left_join(pipedD,by=join.names) %>%
  left_join(sewageN,by=join.names) %>%
  left_join(sewageD,by=join.names) %>%
  left_join(homeownN,by=join.names) %>%
  left_join(homeownD,by=join.names) %>%
  left_join(landownN,by=join.names) %>%
  left_join(landownD,by=join.names) %>%
  left_join(roofmasonN,by=join.names) %>%
  left_join(roofmetalN,by=join.names) %>%  
  left_join(roofwoodN,by=join.names) %>%
  left_join(roofD,by=join.names) %>%
  left_join(trashcollectN,by=join.names) %>%
  left_join(trashcollectD,by=join.names) %>%
  left_join(hhsizeN,by=join.names) %>%
  left_join(hhsizeD,by=join.names)   

c2 <- tot.popn %>%
  left_join(educ,by=join.names) %>%
  left_join(empl,by=join.names) %>%
  left_join(assets, by=join.names) %>%
  left_join(houseatt,by=join.names) %>%
  # to identify countries with ambiguous or missing data for education, employment, and flush toilets
  mutate(edu.detail=ifelse(class(x$EDATTAIND)!="NULL" & all(!is.na(x$EDATTAIND)),1,0),
         empl.detail=ifelse((class(x$EMPSTATD)!="NULL" & all(!is.na(x$EMPSTATD))) & (class(x$OCCISCO)!="NULL" & all(!is.na(x$OCCISCO))),1,0),
         flusht.detail=ifelse("20"%in%unique(x$TOILET),0,1))

#write.csv(c2, 'IPUMS_test_final_output.csv', row.names=FALSE)
   
  sum.table <<- rbind(sum.table,as.data.frame(c2))
# rm(list=setdiff(ls(), c("x","IPUMS_summary","IPUMS_aggr","CSVextract","sum2","sum.table")))
 rm (x)
 # print(paste0("Country ",i, " completed"))
  }
  else {
  rm(x)
 # print(paste0("Country ",i, " missing Geolevel data or numerous age records"))
  } # end if else
}   # end function
```

### IPUMS_aggr function
Loops through all csvs and creates 2 summary tables, one with the sums and the other with the percentages.
**Note** Forced to change Geolevel1 values where there was Geolevel2 data as some records had different Geolevel1 data for the same year and we are grouping by all these values.
```{r eval=FALSE}
IPUMS_aggr <- function(f) {
ptm <- proc.time()               # to figure out how long it takes to run the function
sum.table <<- data.frame()        # create an empty dataframe
  for (k in f){
#print file name
  print(paste0(k,": Country ",gsub("ipums_|.csv","",file.list[k])))
  x <- fread(paste0(loc_files,file.list[k]), showProgress = T)
  IPUMS_summary(x)
  }

# impute the same GEOLEV1 value (using the 1st one) for each GEOLEV2. Creates duplicates otherwise
orig.table <<- sum.table # retain original table with original GEOLEV1 values... just in case
geo2 <-unique(sum.table$GEOLEV2)
geo2 <- subset(geo2,!is.na(geo2))
for (l in geo2) {
    sum.table$GEOLEV1[sum.table$GEOLEV2==l] <- sum.table$GEOLEV1[sum.table$GEOLEV2==l][1]
}
sum.table <- sum.table  %>%
        group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2,edu.detail,empl.detail,flusht.detail) %>%
        summarise_all(sum2)

# calculate percentages
pct.table <- sum.table %>%
  select(COUNTRY:GEOLEV2,edu.detail,empl.detail,flusht.detail,popXX:hhsizeDXX) %>%
  mutate(psecXX=edusecondXX/educDXX,
         pcterXX=educterXX/educDXX,
         plit10XX=lit10NXX/lit10DXX,
         punempXX=unempXX/tlfXX,
         plfpfemXX=lfpfemNXX/lfpfemDXX,
         pprofXX=occprofNXX/occprofDXX,
         pagrXX=occagrNXX/occagrDXX,
         phtwtrXX=hotwaterNXX/hotwaterDXX,
         pcellXX=cellNXX/cellDXX,
         pfrigXX=refrigNXX/refrigDXX,
         ptvXX=tvNXX/tvDXX,
         pradioXX=radioNXX/radioDXX,
         pphoneXX=phoneNXX/phoneDXX,
         pcmptrXX=computerNXX/computerDXX,
         pinternetXX=internetNXX/internetDXX,
         pwashrXX=washerNXX/washerDXX,
         pairconXX=airconNXX/airconDXX,
         pautoXX=autoNXX/autoDXX,
         pelectXX=electricNXX/electricDXX,
         pftoiletXX=flushtoiletNXX/flushtoiletDXX,
         pckelectXX=cookelectNXX/cookDXX,
         pckgasXX=cookgasNXX/cookDXX,
         pckwoodXX=cookwoodNXX/cookDXX,
         ppipedXX=pipedNXX/pipedDXX,
         psewageXX=sewageNXX/sewageDXX,
         phomeXX=homeownNXX/homeownDXX,
         plandXX=landownNXX/landownDXX,
         prfmasonXX=roofmasonNXX/roofDXX,
         prfmetalXX=roofmetalNXX/roofDXX,
         prfwoodXX=roofwoodNXX/roofDXX,
         ptrashXX=trashcollectNXX/trashcollectDXX,
         hhsizeXX=hhsizeNXX/hhsizeDXX) 

#remove original variables
rm.names <- names(sum.table[,-c(1:8)])
pct.table <<- pct.table[,!names(pct.table)%in%rm.names]
sum.table<<-sum.table
proc.time() - ptm
}

```
## Run the functions
Get the list of country csv files in the *loc_files* folder and run the function *IPUMS_aggr* for all files (1:length(file.list)) or for select ones (e.g. (1),(2:10))
```{r eval=FALSE}
IPUMS_aggr(1:length(file.list)) # all files
IPUMS_aggr(44:48) # select files
grep("508",file.list)
#rm(list=setdiff(ls(), c("IPUMS_aggr","IPUMS_summary","sum2","sum.table","pct.table","loc_files","file.list","dsn")))
# show how many NAs in each field
#  apply(x, 2, function(x) paste0(round(sum(is.na(x))/length(x)*100,1),"%"))
```


## Compile GIS and tabular data
**After all the code is fixed and variables chosen**, run this section to organize the data so that there is only 1 row per geolevel unit. Finally, separate the geolevel 1 and 2 data and join to the geolevel shapefiles.
```{r eval=FALSE}
  #-- Geolevel 1 data
glvl1.data <- as.data.frame(pct.table) %>%
  mutate(popXX=popXX/1000) %>%
  filter(is.na(GEOLEV2)) %>%
  gather(var,value, edu.detail:hhsizeXX) %>%
  mutate(var1=paste0(gsub("X","",var),substr(YEAR,3,4))) %>%
  select(-YEAR,-var,-GEOLEV2) %>%
  spread(var1, value) %>%
  rename(GEOLEVEL1=GEOLEV1)%>%
  mutate(GEOLEVEL2=NA) 

glvl1.data$GEOLEVEL1[duplicated(glvl1.data$GEOLEVEL1)] # should be no duplicates

# create decadal means
var.names <-  gsub('XX','',names(pct.table[-c(1:7)]))
for (i in 1:length(var.names)){
#1960s
var.names60 <- grep(paste0(var.names[i],'6'),names(glvl1.data),value = T)
glvl1.data[,paste0(var.names[i],'60s')] <- NA
glvl1.data[,paste0(var.names[i],'60s')] <- ifelse (length(var.names60)>0,
                                                   rowMeans(glvl1.data[var.names60],na.rm = T),
                                                   glvl1.data[,paste0(var.names[i],'60s')])
#1970s
var.names70 <- grep(paste0(var.names[i],'7'),names(glvl1.data),value = T)
glvl1.data[,paste0(var.names[i],'70s')] <- NA
glvl1.data[,paste0(var.names[i],'70s')] <- ifelse(length(var.names70)>0,
                                                   as.data.frame(rowMeans(glvl1.data[var.names70],na.rm = T)),
                                                   glvl1.data[,paste0(var.names[i],'70s')])
#1980s
var.names80 <- grep(paste0(var.names[i],'8'),names(glvl1.data),value = T)
glvl1.data[,paste0(var.names[i],'80s')] <- NA
glvl1.data[,paste0(var.names[i],'80s')] <- ifelse(length(var.names80)>0,
                                                   as.data.frame(rowMeans(glvl1.data[var.names80],na.rm = T)),
                                                   glvl1.data[,paste0(var.names[i],'80s')])
#1990s
var.names90 <- grep(paste0(var.names[i],'9'),names(glvl1.data),value = T)
glvl1.data[,paste0(var.names[i],'90s')] <- NA
glvl1.data[,paste0(var.names[i],'90s')] <- ifelse(length(var.names90)>0,
                                                   as.data.frame(rowMeans(glvl1.data[var.names90],na.rm = T)),
                                                   glvl1.data[,paste0(var.names[i],'90s')])
#2000s
var.names00 <- grep(paste0(var.names[i],'0'),names(glvl1.data),value = T)
glvl1.data[,paste0(var.names[i],'00s')] <- NA
glvl1.data[,paste0(var.names[i],'00s')] <- ifelse(length(var.names00)>0,
                                                   as.data.frame(rowMeans(glvl1.data[var.names00],na.rm = T)),
                                                   glvl1.data[,paste0(var.names[i],'00s')])
#2010s
var.names10 <- grep(paste0(var.names[i],'1'),names(glvl1.data),value = T)
glvl1.data[,paste0(var.names[i],'10s')] <- NA
glvl1.data[,paste0(var.names[i],'10s')] <- ifelse(length(var.names10)>0,
                                                   as.data.frame(rowMeans(glvl1.data[var.names10],na.rm = T)),
                                                   glvl1.data[,paste0(var.names[i],'10s')])
}

  #-- Geolevel 2 data
glvl2.data <- as.data.frame(pct.table) %>%
  mutate(popXX=popXX/100000) %>%
  filter(!is.na(GEOLEV2)) %>%
  gather(var,value, edu.detail:hhsizeXX) %>%
  mutate(var1=paste0(gsub("X","",var),substr(YEAR,3,4)))%>%
  select(-YEAR,-var) %>%
  spread(var1, value) %>%
  rename(GEOLEVEL1=GEOLEV1,
         GEOLEVEL2=GEOLEV2)

glvl2.data$GEOLEVEL2[duplicated(glvl2.data$GEOLEVEL2)]# should be no duplicates

# create decadal means
var.names <-  gsub('XX','',names(pct.table[-c(1:7)]))
for (i in 1:length(var.names)){
#1960s
var.names60 <- grep(paste0(var.names[i],'6'),names(glvl2.data),value = T)
glvl2.data[,paste0(var.names[i],'60s')] <- NA
glvl2.data[,paste0(var.names[i],'60s')] <- ifelse (length(var.names60)>0,
                                                   rowMeans(glvl2.data[var.names60],na.rm = T),
                                                   glvl2.data[,paste0(var.names[i],'60s')])
#1970s
var.names70 <- grep(paste0(var.names[i],'7'),names(glvl2.data),value = T)
glvl2.data[,paste0(var.names[i],'70s')] <- NA
glvl2.data[,paste0(var.names[i],'70s')] <- ifelse(length(var.names70)>0,
                                                   as.data.frame(rowMeans(glvl2.data[var.names70],na.rm = T)),
                                                   glvl2.data[,paste0(var.names[i],'70s')])
#1980s
var.names80 <- grep(paste0(var.names[i],'8'),names(glvl2.data),value = T)
glvl2.data[,paste0(var.names[i],'80s')] <- NA
glvl2.data[,paste0(var.names[i],'80s')] <- ifelse(length(var.names80)>0,
                                                   as.data.frame(rowMeans(glvl2.data[var.names80],na.rm = T)),
                                                   glvl2.data[,paste0(var.names[i],'80s')])
#1990s
var.names90 <- grep(paste0(var.names[i],'9'),names(glvl2.data),value = T)
glvl2.data[,paste0(var.names[i],'90s')] <- NA
glvl2.data[,paste0(var.names[i],'90s')] <- ifelse(length(var.names90)>0,
                                                   as.data.frame(rowMeans(glvl2.data[var.names90],na.rm = T)),
                                                   glvl2.data[,paste0(var.names[i],'90s')])
#2000s
var.names00 <- grep(paste0(var.names[i],'0'),names(glvl2.data),value = T)
glvl2.data[,paste0(var.names[i],'00s')] <- NA
glvl2.data[,paste0(var.names[i],'00s')] <- ifelse(length(var.names00)>0,
                                                   as.data.frame(rowMeans(glvl2.data[var.names00],na.rm = T)),
                                                   glvl2.data[,paste0(var.names[i],'00s')])
#2010s
var.names10 <- grep(paste0(var.names[i],'1'),names(glvl2.data),value = T)
glvl2.data[,paste0(var.names[i],'10s')] <- NA
glvl2.data[,paste0(var.names[i],'10s')] <- ifelse(length(var.names10)>0,
                                                   as.data.frame(rowMeans(glvl2.data[var.names10],na.rm = T)),
                                                   glvl2.data[,paste0(var.names[i],'10s')])
}

```


# Join with spatial data
```{r eval=FALSE}
# Read in aMerge with the shapefiles
Glvl1.shp <- readOGR(dsn,'world_geolev1')
Glvl2.shp <- readOGR(dsn,'world_geolev2')

# create lists of variable names to select and re-order variable names to line up in the shapefiles
# compiling decadal means only
var.names1 <- c("COUNTRY","GEOLEVEL1","GEOLEVEL2",grep('0s',names(glvl1.data),value = T))
var.names2 <- c("COUNTRY","GEOLEVEL1","GEOLEVEL2",grep('0s',names(glvl1.data),value = T))
identical(var.names1,var.names2)  # true?
name.order <- c("GEOLEVEL1","GEOLEVEL2","CNTRY_NAME","ADMIN_NAME","CNTRY_CODE","BPL_CODE",
                grep('0s',names(glvl1.data),value = T))
#names(Glvl1_data.shp)[1:10];names(Glvl2_data.shp)[1:10]
  
# merge each shapefile with the data and merge the 2 shapefiles together
Glvl1_data.shp<- merge(Glvl1.shp,glvl1.data[var.names1],by="GEOLEVEL1") # merge data
Glvl1_data.shp <- Glvl1_data.shp[name.order]
Glvl2_data.shp<- merge(Glvl2.shp,glvl2.data[var.names2],by="GEOLEVEL2") # merge data
Glvl2_data.shp <- Glvl2_data.shp[name.order]
all_data.shp <- rbind(Glvl1_data.shp,Glvl2_data.shp)
writeOGR(all_data.shp,dsn,'IPUMS_geodata',driver = "ESRI Shapefile",overwrite_layer=TRUE,layer_options = "RESIZE=YES")


# output the tabluar data as csv
fwrite(all_data.shp@data,paste0(dsn,"/IPUMS_geodata_decadal_11Nov17.csv"))  # decadal values
fwrite(sum.table,paste0(dsn,"/IPUMS_geodata_sums_11Nov17.csv")) # sums
fwrite(pct.table,paste0(dsn,"/IPUMS_geodata_pct_11Nov17.csv"))  # percentages
fwrite(orig.table,paste0(dsn,"/IPUMS_geodata_orig_11Nov17.csv")) # original sum table with the original GEOL

```


## Data checking
Reading files as a ffdf allows users to work with large dataframes without storing them in memory. # Join with spatial data
```{r eval=FALSE}
# Load shapefile attribute data (decadal means)
ipums.geo <- read.csv(paste0(dsn,"/IPUMS_geodata_decadal_11Nov17.csv"),header = T)

#country codes
ctry.codes <- ipums.geo %>%
  group_by(CNTRY_NAME,CNTRY_CODE) %>%
  summarize(var=n()) %>%
  filter(var>2)

# Load pct table (aggregated to the GEOLVL1/2)
ipums <- read.csv(paste0(dsn,"/IPUMS_geodata_pct_11Nov17.csv"),header = T)
names(ipums) 

# join country name
ipums1 <- ipums %>%
  left_join(ctry.codes[1:2],by=c("COUNTRY"="CNTRY_CODE"))
#----------
# availabilty table
avail.table <- table(ipums1$CNTRY_NAME,ipums$YEAR)
avail.table <- avail.table %>%
  as.data.frame() %>%
  subset(Freq>0) %>%
  rename(Ctry=Var1,Year=Var2,Num=Freq) %>%
  arrange(Ctry,Year)
write.csv(avail.table,"data availability table.csv",row.names = F)

# look at 1 country
Country.check <- ipums1 %>%
  filter(CNTRY_NAME=="Brazil")


hist(pct.table)
#----------
# data availability by variable, country and year
ipums.var <- ipums1 %>%
  select(COUNTRY,CNTRY_NAME,YEAR:GEOLEV2,popXX:hhsizeXX) %>%
  gather(key=variable,value = value,popXX:hhsizeXX) %>%
  filter(!is.na(value)) %>%
  group_by(CNTRY_NAME,YEAR,variable) %>%
  summarise(value=n())  %>%
  unite("ctry_yr",CNTRY_NAME,YEAR) %>%
  arrange(ctry_yr) %>%
  spread(key=ctry_yr,value = value)

write.csv(ipums.var,paste0(dsn,"/IPUMS_variable_summary.csv"),row.names = F)


#----------

  

head(ipums.var)
```


# END