---
title: "IPUMS Data aggregation"
author: "David Gill and Matheus De Nardo"
date: "September 11, 2017"
output: html_document
---

## IPUMS data
This code is used to aggregate social and economic variables from IPUMS at the Geolevel 1 and 2 scale. These data will be used to assess social and economic differences in MPA vs non-MPA communities, both in terms of placement and changes over time. Here data are derived from the IPUMS website on educational attainment and employment, as well as other demographic and economic variables.
### Read in R packages
```{r eval=FALSE}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(data.table)
#library(ff)
#library(ffbase2)
library(rgdal)
options(scipen=999,stringsAsFactors = F)
sum2=function(x){sum(x,na.rm=TRUE)}
mean2=function(x){mean(x,na.rm=TRUE)}
#memory.limit(size = 32000) 

# location of downloaded microdata (country csvs)
loc_files <- 'E:/My Documents/Important Files/Work/SESYNC/Data analysis/IPUMS/data/ipumsi_countries/'
# location of shapefiles
dsn <- "E:/My Documents/Important Files/Work/SESYNC/GIS/IPUMS"# location of shapefiles
```

* Matheus

1. change the file locations above, placing the country csvs in your *loc_files* folder
2. run the two chunks to read in the 2 functions listed below
3. for the section "Compile all regions" run line by line. *IPUMS_aggr(1:length(file.list))* will run all at once, *IPUMS_aggr(1)* will just run it for the 1st csv. Running all of them will take some time (~5 min for 6)
4. try the code all the way to creating the shapefiles. 


## Aggregation functions
There are two functions:

**1. IPUMS_summary**
This function does the following:

* aggregates the data for the select variables of interest at the geolevel2 scale. 
* writes the aggregated data to a new file called **out.table** with the summarized values and **out.table2** with the percentages

**2. IPUMS_aggr**
This function does the following:

* finds and loops through all the files in the folder with the countries
* calculates the percentages

*(need to clean this and place this somewhere else)*
The variables of interest are:
* **Total population**: this gives us total population (based on person weights) within the geography of interest (*popXX*) 
* **Education**: the % of the population who are 25+ years old, for whom educ. attainment is determined, with secondary (*phsXX*) or university (*pcolXX*) level education
* **Unemployment**: the % of individuals within the total workforce (excluding inactive members), for whom employment status is determined, who are unemployed  (*punempXX*).
* **Professional**: the % of individuals within the total employed workforce, for whom employment status is determined, who are employed in a professional occupation  (*pprofXX*). 
* **Female labor force employment**: the % of individuals within the total employed workforce, for whom employment status is determined, who are employed in a professional occupation  (*pflabfXX*).  


### IPUMS_summary function 
summarizes raw values
```{r eval=FALSE}
IPUMS_summary <- function (x) {
  join.names <- c("COUNTRY", "YEAR", "GEOLEV1", "GEOLEV2")
  x <- filter(x,!is.na(GEOLEV1))   # remove entries without Geolevel data (shouldn't happen!)

if (nrow(x)>0 & nrow(filter(x,is.na(AGE)))/nrow(x)<0.1) {  # run if countries are not missing most of the data

tot.popn <- x %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(popXX = sum(PERWT))

#--Education                                    
# number of persons with hs degree or less (Secondary education)
  if (class(x$EDATTAIND)!="NULL" & !is.na(sum2(x$EDATTAIND))) {
educsec <- filter(x,EDATTAIND%in%(311:322) & AGE%in%c(25:100)) %>% # ages 25+ yrs
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(edusecondXX = sum(PERWT))
educD <- filter(x,EDATTAIND%in%(100:400) & AGE%in%c(25:100)) %>% # ages 25+ yrs
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(educDXX = sum(PERWT))

# number persons with at least a four year college degree
educter <- filter(x,EDATTAIND==400 & AGE%in%c(25:100)) %>% # ages 25+ yrs
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(educterXX = sum(PERWT))
   }
  else{
  educsec <- data.frame(tot.popn[,1:4],edusecondXX=NA);educD <- data.frame(tot.popn[,1:4],educDXX=NA);
  educter <- data.frame(tot.popn[,1:4],educterXX=NA);
  }
 # if EDATTAIND is missing but EDATTAIN is present
  if (class(x$EDATTAIN)!="NULL" & !is.na(sum2(x$EDATTAIN))) {
educsec <- filter(x,EDATTAIN%in%(3) & AGE%in%c(25:100)) %>% # ages 25+ yrs
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(edusecondXX = sum(PERWT))
educD <- filter(x,EDATTAIN%in%(1:4) & AGE%in%c(25:100)) %>% # ages 25+ yrs
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(educDXX = sum(PERWT))

# number persons with at least a four year college degree
educter <- filter(x,EDATTAIN==4 & AGE%in%c(25:100)) %>% # ages 25+ yrs
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(educterXX = sum(PERWT))
   }
  else{
  educsec <- educsec;educD <- educD;
  educter <- educter;
  }

# Percent of persons age 10+ literate(numerator)"
  if (class(x$LIT)!="NULL" & !is.na(sum2(x$LIT))) {
lit10N <- filter(x,LIT==2 & AGE%in%c(10:100)) %>% # ages 25+ yrs
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(lit10NXX = sum(PERWT))

# Percent of persons age 10+ literate (denominator)"
lit10D <- filter(x,LIT%in%c(1:2) & AGE%in%c(10:100)) %>% # ages 25+ yrs
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(lit10DXX = sum(PERWT))
   }
  else{
  lit10N <- data.frame(tot.popn[,1:4],lit10NXX=NA);lit10D <- data.frame(tot.popn[,1:4],lit10DXX=NA)
    }

educ <- tot.popn[join.names] %>% 
  left_join(educsec,by = join.names) %>% 
  left_join(educter,by = join.names) %>% 
  left_join(educD,by = join.names) %>% 
  left_join(lit10N,by = join.names) %>% 
  left_join(lit10D,by = join.names)

#--Employment
# Current labor force (employed and unemployed individuals between 16-65; excludes housework)
#empclf <- filter(x,EMPSTATD%in%c(100:240) & AGE%in%c(16:65)) %>%
#  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
#  summarize(empclfXX = sum(PERWT))

# Civilian labor force- persons 16 years and over (excluding armed forces where occupation is known)
  if ((class(x$EMPSTATD)!="NULL" & !is.na(sum2(x$EMPSTATD))) & (class(x$OCCISCO)!="NULL" & !is.na(sum2(x$OCCISCO))) &
      class(x$INDGEN)!="NULL" & !is.na(sum2(x$INDGEN))) {
clf <- filter(x,EMPSTATD%in%c(100:120,140:240) & AGE%in%c(16:65)) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>% #eliminating other column headers
  summarize(clfXX = sum(PERWT)) #can add abbreviated function due to <%> piping
# Unemployment for individuals between 16-65 (of civilians)
unempl <- filter(x,EMPSTATD%in%c(200:240) & AGE%in%c(16:65)) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(unempXX = sum(PERWT)) 

# Females 16 and over, not in armed forces (generally, employed or actively seeking employment)
lfpfemN <- filter(x,SEX==2 & EMPSTATD%in%c(100:120,140) & AGE%in%c(16:65)) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(lfpfemNXX= sum(PERWT))
lfpfemD <- filter(x,SEX==2 & EMPSTATD%in%c(100:120,140:240) & AGE%in%c(16:65)) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(lfpfemDXX= sum(PERWT))

# Percent of workers in professional or technical occupations (excluding clerks)
occprofN <- filter(x,AGE%in%c(16:65) & EMPSTATD%in%c(100:120,140) & OCCISCO%in%c(1:3)) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(occprofNXX = sum(PERWT)) 
occprofD <- filter(x,AGE%in%c(16:65) & EMPSTATD%in%c(100:120,140) & !OCCISCO%in%c(10,97:99)) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(occprofDXX = sum(PERWT)) 

# Percent of currently employed persons age 16 to 65 who are working in agriculture or fisheries
occagrN <- filter(x,AGE%in%c(16:65) & EMPSTATD%in%c(100:120,140) & INDGEN==10) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(occagrNXX = sum(PERWT)) 
occagrD <- filter(x,AGE%in%c(16:65) & EMPSTATD%in%c(100:120,140) & !INDGEN%in%c(0,998:999)) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(occagrDXX = sum(PERWT)) 

empl <- tot.popn[join.names] %>%
  left_join(clf,by=join.names) %>%
  left_join(unempl,by=join.names) %>%
  left_join(lfpfemN,by=join.names) %>%
  left_join(lfpfemD,by=join.names) %>%
  left_join(occprofN,by=join.names) %>%
  left_join(occprofD,by=join.names) %>%
  left_join(occagrN,by=join.names) %>%
  left_join(occagrD,by=join.names) 
   }
  else{
  empl <- data.frame(tot.popn[,1:4],clfXX=NA)
    }


# Civilian labor force- persons 16 years and over (excluding armed forces where occupation is known)
  if (class(x$EMPSTAT)!="NULL" & !is.na(sum2(x$EMPSTAT)) & class(x$OCCISCO)!="NULL" & !is.na(sum2(x$OCCISCO)) &
      class(x$INDGEN)!="NULL" & !is.na(sum2(x$INDGEN))) {
clf <- filter(x,EMPSTAT%in%c(1:2) & AGE%in%c(16:65)) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>% #eliminating other column headers
  summarize(clfXX = sum(PERWT)) #can add abbreviated function due to <%> piping
# Unemployment for individuals between 16-65 (of civilians)
unempl <- filter(x,EMPSTAT%in%c(2) & AGE%in%c(16:65)) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(unempXX = sum(PERWT)) 

# Females 16 and over, not in armed forces (generally, employed or actively seeking employment)
lfpfemN <- filter(x,SEX==2 & EMPSTAT%in%c(1) & AGE%in%c(16:65)) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(lfpfemNXX= sum(PERWT))
lfpfemD <- filter(x,SEX==2 & EMPSTAT%in%c(1:2) & AGE%in%c(16:65)) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(lfpfemDXX= sum(PERWT))

# Percent of workers in professional or technical occupations (excluding clerks)
occprofN <- filter(x,AGE%in%c(16:65) & EMPSTAT%in%c(1) & OCCISCO%in%c(1:3)) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(occprofNXX = sum(PERWT)) 
occprofD <- filter(x,AGE%in%c(16:65) & EMPSTAT%in%c(1) & !OCCISCO%in%c(10,97:99)) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(occprofDXX = sum(PERWT)) 

# Percent of currently employed persons age 16 to 65 who are working in agriculture or fisheries
occagrN <- filter(x,AGE%in%c(16:65) & EMPSTAT%in%c(1) & INDGEN==10) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(occagrNXX = sum(PERWT)) 
occagrD <- filter(x,AGE%in%c(16:65) & EMPSTAT%in%c(1) & !INDGEN%in%c(0,998:999)) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(occagrDXX = sum(PERWT)) 

empl <- tot.popn[join.names] %>%
  left_join(clf,by=join.names) %>%
  left_join(unempl,by=join.names) %>%
  left_join(lfpfemN,by=join.names) %>%
  left_join(lfpfemD,by=join.names) %>%
  left_join(occprofN,by=join.names) %>%
  left_join(occprofD,by=join.names) %>%
  left_join(occagrN,by=join.names) %>%
  left_join(occagrD,by=join.names) 
   }
  else{
  empl <- empl
    }


#--Assets
# Refrigerator ownership
  if (class(x$REFRIG)!="NULL" & !is.na(sum2(x$REFRIG))) {
refrigN <- filter(x,REFRIG==2 & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(refrigNXX = sum(HHWT))
refrigD <- filter(x,REFRIG%in%c(1:2) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(refrigDXX = sum(HHWT)) 
   }
  else{
  refrigN <- data.frame(tot.popn[,1:4],refrigNXX=NA);refrigD <- data.frame(tot.popn[,1:4],refrigDXX=NA)
    }

# hot water access (water heater in unit)
  if (class(x$HOTWATER)!="NULL" & !is.na(sum2(x$HOTWATER))) {
hotwaterN <- filter(x,HOTWATER==2 & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(hotwaterNXX = sum(HHWT))
hotwaterD <- filter(x,HOTWATER%in%c(1:2) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(hotwaterDXX = sum(HHWT)) 
   }
  else{
  hotwaterN <- data.frame(tot.popn[,1:4],hotwaterNXX=NA);hotwaterD <- data.frame(tot.popn[,1:4],hotwaterDXX=NA)
    }

# cellphone access
  if (class(x$CELL)!="NULL" & !is.na(sum2(x$CELL))) {
  cellN <- filter(x,CELL==1 & GQ==10 & PERNUM==1) %>%
        group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
        summarize(cellNXX = sum(HHWT))
  cellD <- filter(x,CELL%in%c(1:2) & GQ==10 & PERNUM==1) %>%
        group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
        summarize(cellDXX = sum(HHWT))
   }
  else{
  cellN <- data.frame(tot.popn[,1:4],cellNXX=NA);cellD <- data.frame(tot.popn[,1:4],cellDXX=NA)
    }


# TV ownership
  if (class(x$TV)!="NULL" & !is.na(sum2(x$TV))) {
tvN <- filter(x,TV%in%c(20:43) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(tvNXX = sum(HHWT))
tvD <- filter(x,TV%in%c(10:43) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(tvDXX = sum(HHWT)) 
   }
  else{
  tvN <- data.frame(tot.popn[,1:4],tvNXX=NA);tvD <- data.frame(tot.popn[,1:4],tvDXX=NA)
    }

# Radio ownership
  if (class(x$RADIO)!="NULL"& !is.na(sum2(x$RADIO))) {
radioN <- filter(x,RADIO==2 & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(radioNXX = sum(HHWT))
radioD <- filter(x,RADIO%in%c(1:2) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(radioDXX = sum(HHWT)) 
   }
  else{
  radioN <- data.frame(tot.popn[,1:4],radioNXX=NA);radioD <- data.frame(tot.popn[,1:4],radioDXX=NA)
    }

# phone ownership
  if (class(x$PHONE)!="NULL" & !is.na(sum2(x$PHONE))) {
phoneN <- filter(x,PHONE==2 & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(phoneNXX = sum(HHWT))
phoneD <- filter(x,PHONE%in%c(1:2) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(phoneDXX = sum(HHWT)) 
   }
  else{
  phoneN <- data.frame(tot.popn[,1:4],phoneNXX=NA);phoneD <- data.frame(tot.popn[,1:4],phoneDXX=NA)
    }

# computer ownership
  if (class(x$COMPUTER)!="NULL" & !is.na(sum2(x$COMPUTER))) {
computerN <- filter(x,COMPUTER==2 & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(computerNXX = sum(HHWT))
computerD <- filter(x,COMPUTER%in%c(1:2) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(computerDXX = sum(HHWT)) 
   }
  else{
  computerN <- data.frame(tot.popn[,1:4],computerNXX=NA);computerD <- data.frame(tot.popn[,1:4],computerDXX=NA)
    }

# internet access
  if (class(x$INTERNET)!="NULL" & !is.na(sum2(x$INTERNET))) {
internetN <- filter(x,INTERNET==2 & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(internetNXX = sum(HHWT))
internetD <- filter(x,INTERNET%in%c(1:2) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(internetDXX = sum(HHWT)) 
   }
  else{
  internetN <- data.frame(tot.popn[,1:4],internetNXX=NA);internetD <- data.frame(tot.popn[,1:4],internetDXX=NA)
    }

# Washer ownership
  if (class(x$WASHER)!="NULL" & !is.na(sum2(x$WASHER))) {
washerN <- filter(x,WASHER%in%c(2:4) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(washerNXX = sum(HHWT))
washerD <- filter(x,WASHER%in%c(1:4) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(washerDXX = sum(HHWT)) 
   }
  else{
  washerN <- data.frame(tot.popn[,1:4],washerNXX=NA);washerD <- data.frame(tot.popn[,1:4],washerDXX=NA)
    }

# Air-conditioner ownership
  if (class(x$AIRCON)!="NULL" & !is.na(sum2(x$AIRCON))) {
airconN <- filter(x,AIRCON%in%c(20:29) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(airconNXX = sum(HHWT))
airconD <- filter(x,AIRCON%in%c(10:29) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(airconDXX = sum(HHWT)) 
   }
  else{
  airconN <- data.frame(tot.popn[,1:4],airconNXX=NA);airconD <- data.frame(tot.popn[,1:4],airconDXX=NA)
    }

# Auto ownership
  if (class(x$AUTOS)!="NULL" & !is.na(sum2(x$AUTOS))) {
autoN <- filter(x,AUTOS%in%c(1:7) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(autoNXX = sum(HHWT))
autoD <- filter(x,AUTOS%in%c(0:7) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(autoDXX = sum(HHWT)) 
   }
  else{
  autoN <- data.frame(tot.popn[,1:4],autoNXX=NA);autoD <- data.frame(tot.popn[,1:4],autoDXX=NA)
    }


assets <- tot.popn[join.names]   %>%
  left_join(hotwaterN,by=join.names) %>%
  left_join(hotwaterD,by=join.names) %>%
  left_join(cellN,by=join.names) %>%
  left_join(cellD,by=join.names) %>%
  left_join(refrigN,by=join.names) %>%
  left_join(refrigD,by=join.names) %>%
  left_join(tvN,by=join.names) %>%
  left_join(tvD,by=join.names) %>% 
  left_join(radioN,by=join.names) %>%
  left_join(radioD,by=join.names) %>%
  left_join(phoneN,by=join.names) %>%
  left_join(phoneD,by=join.names) %>%
  left_join(computerN,by=join.names) %>%
  left_join(computerD,by=join.names) %>%
  left_join(internetN,by=join.names) %>%
  left_join(internetD,by=join.names) %>%
  left_join(washerN,by=join.names) %>%
  left_join(washerD,by=join.names) %>%
  left_join(airconN,by=join.names) %>%
  left_join(airconD,by=join.names) %>%
  left_join(autoN,by=join.names) %>%
  left_join(autoD,by=join.names)


#-- Household attributes
#Percent of households with electricity
  if (class(x$ELECTRIC)!="NULL" & !is.na(sum2(x$ELECTRIC))) {
electricN <- filter(x,ELECTRIC==1 & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(electricNXX = sum(HHWT))
electricD <- filter(x,ELECTRIC%in%c(1:2) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(electricDXX = sum(HHWT)) 
   }
  else{
  electricN <- data.frame(tot.popn[,1:4],electricNXX=NA);electricD <- data.frame(tot.popn[,1:4],electricDXX=NA)
    }

#Percent of households with a flush toilet
  if (class(x$TOILET)!="NULL" & !is.na(sum2(x$TOILET))) {
flushtoiletN <- filter(x,TOILET==21 & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize( flushtoiletNXX = sum(HHWT))
flushtoiletD <- filter(x,TOILET%in%c(10:23) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(flushtoiletDXX = sum(HHWT))
   }
  else{
  flushtoiletN <- data.frame(tot.popn[,1:4],flushtoiletNXX=NA);flushtoiletD <- data.frame(tot.popn[,1:4],flushtoiletDXX=NA)
    }

# Cooking fuel
#Electricity
  if (class(x$FUELCOOK)!="NULL" & !is.na(sum2(x$FUELCOOK))) {
cookelectN <- filter(x,FUELCOOK==20 & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(cookelectNXX = sum(HHWT))
cookD <- filter(x,FUELCOOK%in%c(10:77) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(cookDXX = sum(HHWT))
#Gas
cookgasN <- filter(x,FUELCOOK%in%c(30:47) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(cookgasNXX = sum(HHWT))
#Wood and solid fuels
cookwoodN <- filter(x,FUELCOOK%in%c(50:56) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(cookwoodNXX = sum(HHWT))
   }
  else{
  cookelectN <- data.frame(tot.popn[,1:4],cookelectNXX=NA); cookD <- data.frame(tot.popn[,1:4],cookDXX=NA);
  cookgasN <- data.frame(tot.popn[,1:4],cookgasNXX=NA); cookwoodN <- data.frame(tot.popn[,1:4],cookwoodNXX=NA)
    }

#Percentage of households with access to piped water
  if (class(x$WATSUP)!="NULL" & !is.na(sum2(x$WATSUP))) {
pipedN <- filter(x,WATSUP%in%c(10:18) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(pipedNXX = sum(HHWT))
pipedD <- filter(x,WATSUP%in%c(10:20) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(pipedDXX = sum(HHWT))
   }
  else{
  pipedN <- data.frame(tot.popn[,1:4],pipedNXX=NA);pipedD <- data.frame(tot.popn[,1:4],pipedDXX=NA)
    }

#Percentage of households with access to sewage facilities
  if (class(x$SEWAGE)!="NULL" & !is.na(sum2(x$SEWAGE))) {
sewageN <- filter(x,SEWAGE%in%c(10:12) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(sewageNXX = sum(HHWT))
sewageD <- filter(x,SEWAGE%in%c(10:20) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(sewageDXX = sum(HHWT))
   }
  else{
  sewageN <- data.frame(tot.popn[,1:4],sewageNXX=NA);sewageD <- data.frame(tot.popn[,1:4],sewageDXX=NA)
    }

#Percentage household ownership
  if (class(x$OWNERSHIP)!="NULL" & !is.na(sum2(x$OWNERSHIP))) {
homeownN <- filter(x,OWNERSHIP==1 & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(homeownNXX = sum(HHWT))
homeownD <- filter(x,OWNERSHIP%in%c(1:2) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(homeownDXX = sum(HHWT))
   }
  else{
  homeownN <- data.frame(tot.popn[,1:4],homeownNXX=NA);homeownD <- data.frame(tot.popn[,1:4],homeownDXX=NA)
    }

#Percentage land ownership
  if (class(x$LANDOWN)!="NULL" & !is.na(sum2(x$LANDOWN))) {
landownN <- filter(x,LANDOWN%in%c(10:14) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(landownNXX = sum(HHWT))
landownD <- filter(x,LANDOWN%in%c(10:30) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(landownDXX = sum(HHWT))
   }
  else{
  landownN <- data.frame(tot.popn[,1:4],landownNXX=NA);landownD <- data.frame(tot.popn[,1:4],landownDXX=NA)
    }

#Roof types
# Masonary, concrete, tiles
  if (class(x$ROOF)!="NULL" & !is.na(sum2(x$ROOF))) {
roofmasonN <- filter(x,ROOF%in%c(10:29) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(roofmasonNXX = sum(HHWT))
roofD <- filter(x,ROOF%in%c(10:90) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(roofDXX = sum(HHWT))
# Metal
roofmetalN <- filter(x,ROOF%in%c(30:35) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(roofmetalNXX = sum(HHWT))
# Wood and/or clay
roofwoodN <- filter(x,ROOF%in%c(40:61) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(roofwoodNXX = sum(HHWT))
   }
  else{
  roofmasonN <- data.frame(tot.popn[,1:4],roofmasonNXX=NA);roofD <- data.frame(tot.popn[,1:4],roofDXX=NA);
  roofmetalN <- data.frame(tot.popn[,1:4],roofmetalNXX=NA);roofwoodN <- data.frame(tot.popn[,1:4],roofwoodNXX=NA)
    }

#trash
# Collected or not collected
  if (class(x$TRASH)!="NULL") {
trashcollectN <- filter(x,TRASH%in%c(10:14) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(trashcollectNXX = sum(HHWT))
trashcollectD <- filter(x,TRASH%in%c(10:39) & GQ==10 & PERNUM==1) %>%
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(trashcollectDXX = sum(HHWT))
   }
  else{
  trashcollectN <- data.frame(tot.popn[,1:4],trashcollectNXX=NA);trashcollectD <- data.frame(tot.popn[,1:4],trashcollectDXX=NA)
    }

#Person per household                                      
hhsize <- filter(x, GQ==10 & PERNUM==1) %>% 
  group_by(COUNTRY,YEAR,GEOLEV1,GEOLEV2) %>%
  summarize(hhsizeXX = sum2(PERSONS*HHWT)/sum2(HHWT))


houseatt <- tot.popn[join.names]   %>%
  left_join(electricN,by=join.names) %>%
  left_join(electricD,by=join.names) %>%
  left_join(flushtoiletN,by=join.names) %>%
  left_join(flushtoiletD,by=join.names) %>%
  left_join(cookelectN,by=join.names) %>%
  left_join(cookgasN,by=join.names) %>%
  left_join(cookwoodN,by=join.names) %>%
  left_join(cookD,by=join.names) %>% 
  left_join(pipedN,by=join.names) %>%
  left_join(pipedD,by=join.names) %>%
  left_join(sewageN,by=join.names) %>%
  left_join(sewageD,by=join.names) %>%
  left_join(homeownN,by=join.names) %>%
  left_join(homeownD,by=join.names) %>%
  left_join(landownN,by=join.names) %>%
  left_join(landownD,by=join.names) %>%
  left_join(roofmasonN,by=join.names) %>%
  left_join(roofmetalN,by=join.names) %>%  
  left_join(roofwoodN,by=join.names) %>%
  left_join(roofD,by=join.names) %>%
  left_join(trashcollectN,by=join.names) %>%
  left_join(trashcollectD,by=join.names) %>%
  left_join(hhsize,by=join.names)
  

c2 <- tot.popn %>%
  left_join(educ,by=join.names) %>%
  left_join(empl,by=join.names) %>%
  left_join(assets, by=join.names) %>%
  left_join(houseatt,by=join.names) 

#write.csv(c2, 'IPUMS_test_final_output.csv', row.names=FALSE)
   
  out.table <<- rbind(out.table,as.data.frame(c2))
# rm(list=setdiff(ls(), c("x","IPUMS_summary","IPUMS_aggr","CSVextract","sum2","out.table")))
 rm (x)
 # print(paste0("Country ",i, " completed"))
  }
  else {
  rm(x)
 # print(paste0("Country ",i, " missing Geolevel data or numerous age records"))
  } # end if else
}   # end function
```

### IPUMS_aggr function
Loops through all csvs and creates 2 summary tables, one with the sums and the other with the percentages
```{r eval=FALSE}
IPUMS_aggr <- function(f) {
ptm <- proc.time()               # to figure out how long it takes to run the function
file.list<- list.files(loc_files,recursive = T, full.names = F) # get list of files in folder
out.table <<- data.frame()        # create an empty dataframe
  for (k in f){
#print file name
  print(file.list[k])
  x <- fread(paste0(loc_files,file.list[k]), showProgress = T)
  IPUMS_summary(x)
  }
out.table2 <- out.table %>%
  mutate(psecXX=edusecondXX/educDXX,
         pcterXX=educterXX/educDXX,
         plit10XX=lit10NXX/lit10DXX,
         punempXX=unempXX/clfXX,
         plfpfemXX=lfpfemNXX/lfpfemDXX,
         pprofXX=occprofNXX/occprofDXX,
         pagrXX=occagrNXX/occagrDXX,
         phtwtrXX=hotwaterNXX/hotwaterDXX,
         pcellXX=cellNXX/cellDXX,
         pfrigXX=refrigNXX/refrigDXX,
         ptvXX=tvNXX/tvDXX,
         pradioXX=radioNXX/radioDXX,
         pphoneXX=phoneNXX/phoneDXX,
         pcmptrXX=computerNXX/computerDXX,
         pinternetXX=internetNXX/internetDXX,
         pwashrXX=washerNXX/washerDXX,
         pairconXX=airconNXX/airconDXX,
         pautoXX=autoNXX/autoDXX,
         pelectXX=electricNXX/electricDXX,
         pftoiletXX=flushtoiletNXX/flushtoiletDXX,
         pckelectXX=cookelectNXX/cookDXX,
         pckgasXX=cookgasNXX/cookDXX,
         pckwoodXX=cookwoodNXX/cookDXX,
         ppipedXX=pipedNXX/pipedDXX,
         psewageXX=sewageNXX/sewageDXX,
         phomeXX=homeownNXX/homeownDXX,
         plandXX=landownNXX/landownDXX,
         prfmasonXX=roofmasonNXX/roofDXX,
         prfmetalXX=roofmetalNXX/roofDXX,
         prfwoodXX=roofwoodNXX/roofDXX,
         ptrashXX=trashcollectNXX/trashcollectDXX)

#remove original variables
rm.names <- names(out.table[,-c(1:5,63)])
out.table2 <<- out.table2[,!names(out.table2)%in%rm.names]
names(out.table2)

#rm(list=setdiff(ls(), c("IPUMSagg","IPUMS_summary","sum2","out.table","out.table2","loc_files","file.list","dsn")))
proc.time() - ptm
}
  # show how many NAs in each field
#  apply(x, 2, function(x) paste0(round(sum(is.na(x))/length(x)*100,1),"%"))
```

## Compile all regions
**After all the code is fixed and variables chosen**, run this section to organize the data so that there is only 1 row per geolevel unit. Finally, separate the geolevel 1 and 2 data and join to the geolevel shapefiles.
```{r eval=FALSE}

IPUMS_aggr(1:length(file.list)) # all files
IPUMS_aggr(1) # just the 1st file

#-- Geolevel 1 data
glvl1.data <- out.table2 %>%
  filter(is.na(GEOLEV2)) %>%
  gather(var,value, popXX:ptrashXX) %>%
  mutate(var1=paste0(gsub("X","",var),substr(YEAR,3,4)))%>%
  select(-YEAR,-var,-GEOLEV2) %>%
  spread(var1, value) %>%
  rename(GEOLEVEL1=GEOLEV1)%>%
  mutate(GEOLEVEL2=NA)


# create decadal means
var.names <-  gsub('XX','',names(out.table2[-c(1:4)]))
for (i in 1:length(var.names)){
#1960s
var.names60 <- grep(paste0(var.names[i],'6'),names(glvl1.data),value = T)
glvl1.data[,paste0(var.names[i],'60s')] <- NA
glvl1.data[,paste0(var.names[i],'60s')] <- ifelse (length(var.names60)>0,
                                                   rowMeans(glvl1.data[var.names60],na.rm = T),
                                                   glvl1.data[,paste0(var.names[i],'60s')])
#1970s
var.names70 <- grep(paste0(var.names[i],'7'),names(glvl1.data),value = T)
glvl1.data[,paste0(var.names[i],'70s')] <- NA
glvl1.data[,paste0(var.names[i],'70s')] <- ifelse(length(var.names70)>0,
                                                   as.data.frame(rowMeans(glvl1.data[var.names70],na.rm = T)),
                                                   glvl1.data[,paste0(var.names[i],'70s')])
#1980s
var.names80 <- grep(paste0(var.names[i],'8'),names(glvl1.data),value = T)
glvl1.data[,paste0(var.names[i],'80s')] <- NA
glvl1.data[,paste0(var.names[i],'80s')] <- ifelse(length(var.names80)>0,
                                                   as.data.frame(rowMeans(glvl1.data[var.names80],na.rm = T)),
                                                   glvl1.data[,paste0(var.names[i],'80s')])
#1990s
var.names90 <- grep(paste0(var.names[i],'9'),names(glvl1.data),value = T)
glvl1.data[,paste0(var.names[i],'90s')] <- NA
glvl1.data[,paste0(var.names[i],'90s')] <- ifelse(length(var.names90)>0,
                                                   as.data.frame(rowMeans(glvl1.data[var.names90],na.rm = T)),
                                                   glvl1.data[,paste0(var.names[i],'90s')])
#2000s
var.names00 <- grep(paste0(var.names[i],'0'),names(glvl1.data),value = T)
glvl1.data[,paste0(var.names[i],'00s')] <- NA
glvl1.data[,paste0(var.names[i],'00s')] <- ifelse(length(var.names00)>0,
                                                   as.data.frame(rowMeans(glvl1.data[var.names00],na.rm = T)),
                                                   glvl1.data[,paste0(var.names[i],'00s')])
#2010s
var.names10 <- grep(paste0(var.names[i],'1'),names(glvl1.data),value = T)
glvl1.data[,paste0(var.names[i],'10s')] <- NA
glvl1.data[,paste0(var.names[i],'10s')] <- ifelse(length(var.names10)>0,
                                                   as.data.frame(rowMeans(glvl1.data[var.names10],na.rm = T)),
                                                   glvl1.data[,paste0(var.names[i],'10s')])
}


#-- Geolevel 2 data
glvl2.data <- out.table2 %>%
  filter(!is.na(GEOLEV2)) %>%
  gather(var,value, popXX:ptrashXX) %>%
  mutate(var1=paste0(gsub("X","",var),substr(YEAR,3,4)))%>%
  select(-YEAR,-var) %>%
  spread(var1, value) %>%
  rename(GEOLEVEL1=GEOLEV1,
         GEOLEVEL2=GEOLEV2)

# create decadal means
var.names <-  gsub('XX','',names(out.table2[-c(1:4)]))
for (i in 1:length(var.names)){
#1960s
var.names60 <- grep(paste0(var.names[i],'6'),names(glvl2.data),value = T)
glvl2.data[,paste0(var.names[i],'60s')] <- NA
glvl2.data[,paste0(var.names[i],'60s')] <- ifelse (length(var.names60)>0,
                                                   rowMeans(glvl2.data[var.names60],na.rm = T),
                                                   glvl2.data[,paste0(var.names[i],'60s')])
#1970s
var.names70 <- grep(paste0(var.names[i],'7'),names(glvl2.data),value = T)
glvl2.data[,paste0(var.names[i],'70s')] <- NA
glvl2.data[,paste0(var.names[i],'70s')] <- ifelse(length(var.names70)>0,
                                                   as.data.frame(rowMeans(glvl2.data[var.names70],na.rm = T)),
                                                   glvl2.data[,paste0(var.names[i],'70s')])
#1980s
var.names80 <- grep(paste0(var.names[i],'8'),names(glvl2.data),value = T)
glvl2.data[,paste0(var.names[i],'80s')] <- NA
glvl2.data[,paste0(var.names[i],'80s')] <- ifelse(length(var.names80)>0,
                                                   as.data.frame(rowMeans(glvl2.data[var.names80],na.rm = T)),
                                                   glvl2.data[,paste0(var.names[i],'80s')])
#1990s
var.names90 <- grep(paste0(var.names[i],'9'),names(glvl2.data),value = T)
glvl2.data[,paste0(var.names[i],'90s')] <- NA
glvl2.data[,paste0(var.names[i],'90s')] <- ifelse(length(var.names90)>0,
                                                   as.data.frame(rowMeans(glvl2.data[var.names90],na.rm = T)),
                                                   glvl2.data[,paste0(var.names[i],'90s')])
#2000s
var.names00 <- grep(paste0(var.names[i],'0'),names(glvl2.data),value = T)
glvl2.data[,paste0(var.names[i],'00s')] <- NA
glvl2.data[,paste0(var.names[i],'00s')] <- ifelse(length(var.names00)>0,
                                                   as.data.frame(rowMeans(glvl2.data[var.names00],na.rm = T)),
                                                   glvl2.data[,paste0(var.names[i],'00s')])
#2010s
var.names10 <- grep(paste0(var.names[i],'1'),names(glvl2.data),value = T)
glvl2.data[,paste0(var.names[i],'10s')] <- NA
glvl2.data[,paste0(var.names[i],'10s')] <- ifelse(length(var.names10)>0,
                                                   as.data.frame(rowMeans(glvl2.data[var.names10],na.rm = T)),
                                                   glvl2.data[,paste0(var.names[i],'10s')])
}
```


# Join with spatial data
```{r eval=FALSE}
names(Glvl1_data.shp)[1:10];names(Glvl2_data.shp)[1:10]
# Read in aMerge with the shapefiles
Glvl1.shp <- readOGR(dsn,'world_geolev1')
Glvl2.shp <- readOGR(dsn,'world_geolev2')

# create lists of variable names to select and re-order variable names to line up in the shapefiles
var.names1 <- c("COUNTRY","GEOLEVEL1","GEOLEVEL2",grep('0s',names(glvl1.data),value = T))
var.names2 <- c("COUNTRY","GEOLEVEL1","GEOLEVEL2",grep('0s',names(glvl1.data),value = T))
name.order <- c("GEOLEVEL1","GEOLEVEL2","CNTRY_NAME","ADMIN_NAME","CNTRY_CODE","BPL_CODE",grep('0s',names(glvl1.data),value = T))
  
# merge each shapefile with the data and merge the 2 shapefiles together
Glvl1_data.shp<- merge(Glvl1.shp,glvl1.data[var.names1],by="GEOLEVEL1") # merge data
Glvl1_data.shp <- Glvl1_data.shp[name.order]
Glvl2_data.shp<- merge(Glvl2.shp,glvl2.data[var.names2],by="GEOLEVEL2") # merge data
Glvl2_data.shp <- Glvl2_data.shp[name.order]
all_data.shp <- rbind(Glvl1_data.shp,Glvl2_data.shp)
writeOGR(all_data.shp,dsn,'IPUMS_geodata',driver = "ESRI Shapefile")
```


# END

## Read in the data in ffdf format
Reading files as a ffdf allows users to work with large dataframes without storing them in memory. Although they take quite a while to read into R and to convert to dataframes, they can be queried (i.e. you can select portions of the data you need) using the *filter* function.
```{r eval=FALSE}
#system.time(asia<- read.csv.ffdf(file="ipumsi_asia.csv", header=TRUE, VERBOSE=TRUE, 
#                  first.rows=1000000, next.rows=1000000, colClasses=NA))
#system.time(x <- fread('ipumsi_00028.csv',nrows=10000000))
#headernames <- names(x)
#pryr::object_size(asia)
system.time(x <- fread('data/ipumsi_00049.csv/ipumsi_00049.csv', showProgress = T))

```

